Equalizer\\Programming and User Guide

### 2011-02-22 Developer/ProgrammingGuide/paper.tex
Programming and User Guide Update.
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.10 for Equalizer 1.0$\alpha$ \\[\medskipamount]
%  \today
}

@@

\lstset{language=C++}
\section*{Equalizer 1.0$\alpha$ Programming and User Guide}
\today
\subsection*{Contributors}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
    1.7.7   & Oct 01, 2010 & SystemPipe and SystemWindow renaming\\
    1.7.8   & Oct 04, 2010 & Document GPU affinity workaround hint\\
    1.7.9   & Nov 03, 2010 & Add \sref{sErrorHandling}\\
    1.7.10  & Dec 21, 2010 & Review for Equalizer 1.0$\alpha$\\
    \hline
  \end{tabularx}}
\clearpage

@@

For non-planar displays, configure the frustum as described in
\sref{sCanvas}. For passive stereo installations, configure one segment
per eye pass, where the segment for the left and right eye have the same
viewport.

@@

Configure one \textsf{view} for each logical view in each layout. Set
the \textsf{viewport} to position the view.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/syncAsync}
  {\caption{\label{fSyncAsync}Synchronous and Asynchronous Execution}}
\end{figure}
\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
2D compound without latency and with a latency of one frame. The
asynchronous execution pipelines certain rendering operations and hides
imbalances in the load distribution, resulting in an improved
framerate. For example, we have observed a speedup of 15\% on a
five-node rendering cluster when using a latency of one frame instead of
no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

{\footnotesize\lstinputlisting[firstline=122, lastline=129]
  {../../../src/examples/eqPly/frameData.h}}
{\footnotesize\lstinputlisting[firstline=51, lastline=77]
  {../../../src/examples/eqPly/frameData.cpp}}

@@

\begin{figure}[ht!]\center
%\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}
  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
  {\caption{\label{fConfigInit}Config Initialization Sequence}}
  \vspace{-4ex}
\end{figure}
The tasks are executed by the node and pipe threads in parallel. The parent's
initialization methods are always executed before any child initialization
method. This parallelization is necessary to allow a speedy startup of the
configuration on large-scale graphics clusters. On the other hand, it means that
initialization functions are called even if the parent's initialization has
failed. \fig{fConfigInit} shows a sequence diagram of the config initialization.

@@

{\footnotesize\lstinputlisting[firstline=314, lastline=321]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=323, lastline=331]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=333, lastline=363]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=391, lastline=394]
  {../../../src/examples/eqPly/config.cpp}}
{\footnotesize\lstinputlisting[firstline=454, lastline=485]
  {../../../src/examples/eqPly/config.cpp}}
{\footnotesize\lstinputlisting[firstline=532, lastline=538]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=288, lastline=312]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=833, lastline=870]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=261, lastline=272]
  {../../../src/examples/eqPly/config.cpp}}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\label{fDBDest}Destination View of a DB Compound using
      Demonstrative Coloring}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the frame data to ensure
that all channels render a given frame using the same position.

@@

{\footnotesize\lstinputlisting[firstline=600, lastline=621]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=628, lastline=698]
  {../../../src/examples/eqPly/channel.cpp}}

@@

The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
\end{wrapfigure}
For each window, one \textsf{GLXEventHandler} is allocated. FBO's and PBuffers
are handled in the same way as window drawables. The event dispatch finds the
corresponding event handler for each received event, which is then used to
process the event in a fashion similar to AGL. The \textsf{GLXWindowEvent}
passes the \textsf{XEvent} to \textsf{GLXWindowIF::\-process\-Event}.

@@

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all their children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame}
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
\fig{fFrameSync} illustrates the local frame synchronization.

@@

GLX initialization is very similar to AGL initialization. Again the
steps are: choose visual (pixel format), create OpenGL context and then
create drawable. The only difference is that the data returned by
\textsf{chooseXVisualInfo} has to be freed using \textsf{XFree}:

{\footnotesize\lstinputlisting[firstline=67, lastline=95]
  {../../../src/libs/client/glXWindow.cpp}}

@@

The render clients can access a slave instance of the view using
\textsf{Channel::getView}. When called from one of the frame task
methods, this method will return the view of the current destination
channel for which the task method is executed. Otherwise it returns the
channel's native view, if it has one. Only destination channels of an
active canvas have a native view.

@@

\begin{wrapfigure}{r}{.612\textwidth}
  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
\end{wrapfigure}
The most common entity to subclass is the \textsf{View}, since the application
often amends with view-specific application data. A view might have a
distributable user data object, which has to inherit from \textsf{co::Object} to
be committed and synchronized with the associated object.

@@

\subsubsection{Dynamic Layout Switch}

@@

@@

A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
application process can set the wall or projection parameters at
runtime. For a description of wall and projection parameters please refer
to \sref{sFrustum}. The new data will be effective for the next frame.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
\end{wrapfigure}
The frustum of a view overrides the underlying frustum of the
segments. The frustum of a view does not typically correspond to a
physical projection, and can therefore be changed at runtime. The
frustum of a segment does however correspond to a physical projection,
and is therefore immutable.

@@

\begin{wrapfigure}{l}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

Finding the correct assembly order for perspective frusta is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frusta. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.

@@

The eqPly example contains support for software anti-aliasing when being idle,
showing how to combine it with subpixel compounds. The same basic algorithm
might be applied to ohter multi-sampling effects, e.g., depth-of-field.

\subsubsection{Supersampling Algorithm}

@@

The subpixel area is a function of the current jitter step and the channel's
\textsf{subpixel} description. To evenly distribute jitter values for all
channels over the total sub-pixel grid a precomputed prime number table is used
to pseudo-randomly select the subpixel:

{\footnotesize\lstinputlisting[firstline=556, lastline=579]
  {../../../src/examples/eqPly/channel.cpp}}

@@

During idle rendering of any decomposition, the jitter for the frustum is
computed using the normalized subpixel center point and the size of a pixel on
the near plane. A random position within the sub-pixel is set as a sample
position, which will be used to move the frustum:

{\footnotesize\lstinputlisting[firstline=512, lastline=554]
  {../../../src/examples/eqPly/channel.cpp}}

\subsubsection{Idle Anti-Aliasing in eqPly}

@@

If the application is in idle mode, the \textsf{applyFrustum} operation will 
apply the computed jitter vector to the current frustum:

{\footnotesize\lstinputlisting[firstline=377, lastline=400]
  {../../../src/examples/eqPly/channel.cpp}}

@@

Statistics are measured in milliseconds since the configuration was
initialized. On each node, the global configuration clock is synchronized
with the server clock. Each statistic event records the originator's
(channel, window, frame or config) unique identifier.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/statistics}
  {\caption{\label{fStatistics}Statistics for a two-Node 2D Compound}}
\end{wrapfigure}
Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

@@

\subsubsection{Statistics Overlay}

The Equalizer examples render a statistics overlay using the gathered
statistics events directly before
\textsf{Window::swapBuffers}. Statistics are toggled on and off using
the 's' key. The overlay rendering method,
\textsf{Channel::drawStatistics}, can be called directly by any
application to render the overlay. It can also be overwritten by a
custom implementation.

\fig{fStatisticsDetail} shows a detail view of \fig{fStatistics}. The
statistics shown are for a two-node 2D compound. The destination channel
is on the \textsf{appNode} and contributes to the rendering.

@@

The X axis is the time. One pixel on the screen corresponds to a given
time unit, here one millisecond per pixel. The scale is zoomed dynamically
to powers-of-ten milliseconds to fit the statistics into the available
viewport. This allows easy and accurate evaluations of bottlenecks or
misconfigurations in the rendering pipeline. The scale of the statistics
is printed.

The rightmost pixel is the most current time. On the Y axis are the
entities: channels, windows, frames and the config. The upper channel is
the local channel since it executes \textsf{frameAssemble}, and the
lower channel is the remote channel, executing \textsf{frameReadback}.

The draw time has been artificially inflated by using immediate mode
rendering to provide more expressive statistics.

To facilitate the understanding, older frames are gradually grayed
out. The rightmost, current frame is brighter than the
frame-before-last.

The configuration used has a latency of one frame. Consequently, the
execution of two frames overlaps. This can be observed in the early
execution of the remote channel's \textsf{frameDraw}, which starts while
the local channel is still assembling the previous frame.

The beginning of a frame is marked by a vertical green line, and the end
of a frame by a vertical gray line. These lines are also attenuated. The
brightness and color matches the event for \textsf{Config::startFrame}
and \textsf{Config::finishFrame}, respectively. The event for
\textsf{startFrame} is typically not visible, since it takes less than
one millisecond to execute. If no idle processing is done by the
application, the event for \textsf{finishFrame} occupies a full frame,
since the config is blocked here waiting for the frame to complete.

In the above example, the local channel finishes drawing the frame
early, and therefore spends a considerable amount of time waiting for
the input frame from the remote channel. These wait events, rendered
red, are a sub-event of the yellow \textsf{frameAssemble} task.

The frame statistics provide information on the image transmission for
scalable rendering. They are grouped per frame, not per node, that is,
the transmit and compress part are executed by one node and the receive
by another, even though the statistics are drawn in the same line. The
decompression time is part of the receive statistic. The total
compression ratio is printed with each compression statistic, in this
case the image has been compressed to 11\% of its original size.

The white \textsf{Window::swapBuffers} task might also take also a long
time, since the execution of the swap buffer is locked to the vertical
retrace of the display, and enforced by \textsf{glFinish} before
sampling the end time of the event. In release builds, this does not
happen and all OpenGL commands are pipelined with the execution on the
CPU.

Note that the remote source window does not execute the swapBuffers in
this configuration, since it is a single-buffered PBuffer.

@@

\fig{fNetNode} shows the communication between two nodes. The nodes in
the listening state also have a listening connection, which is not shown
in the Figure for complexity reasons.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
  {\caption{\label{fNetNode}Communication between two Nodes}}
\end{wrapfigure}
When the remote node sends a packet, the listening node receives the
packet and dispatches it from the receiver thread using the method
\textsf{dispatchCommand}. The default implementation knows how to
dispatch packets of type node, session or object. Applications can
define custom data types for packets, and then have to extend
\textsf{dispatchCommand} to handle these custom data types.

@@

Multiple instances of the same object (identifier) can be mapped on the same
node and session simultaneously. Each object instance has a node-unique instance
identifier. When the instance ID is set to \textsf{EQ\_INSTANCE\_ALL} (the
default value), a command is dispatch to all object instances registered to the
session with the correct object ID. Multiple object instance per node are used
when different threads on the node need to process a different version of the
same object.

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.8.RC1 for Equalizer 1.0$\beta$ \\[\medskipamount]
%  \today
}

@@

\lstset{language=C++}
\section*{Equalizer 1.0$\beta$ Programming and User Guide}
\today
\subsection*{Contributors}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
    1.7.7   & Oct 01, 2010 & SystemPipe and SystemWindow renaming\\
    1.7.8   & Oct 04, 2010 & Document GPU affinity workaround hint\\
    1.7.9   & Nov 03, 2010 & Add \sref{sErrorHandling}\\
    1.7.10  & Dec 21, 2010 & Review for Equalizer 1.0$\alpha$\\
    1.8.RC1  & Feb 22, 2010 & Review for Equalizer 1.0$\beta$\\
    \hline
  \end{tabularx}}
\clearpage

@@

For non-planar displays, configure the frustum as described in
\sref{sCanvas}. For passive stereo installations, configure one segment per eye
pass, where the segment for the left and right eye have the same viewport. Set
the eyes displayed by the segment, i.e., left or right and potentially cyclop.

@@

Configure one \textsf{view} for each logical view in each layout. Set the
\textsf{viewport} to position the view. Set the mode to stereo if appropriate.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/syncAsync}
  {\caption{\label{fSyncAsync}Synchronous and Asynchronous Execution}}
\end{figure}
\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
2D compound without latency and with a latency of one frame. The
asynchronous execution pipelines certain rendering operations and hides
imbalances in the load distribution, resulting in an improved
framerate. For example, we have observed a speedup of 15\% on a
five-node rendering cluster when using a latency of one frame instead of
no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user. The
statistics overlay is explained in detail in \sref{sStatisticsOverlay}.

@@

{\footnotesize\lstinputlisting[firstline=122, lastline=129]
  {../../../src/examples/eqPly/frameData.h}}
{\footnotesize\lstinputlisting[firstline=54, lastline=80]
  {../../../src/examples/eqPly/frameData.cpp}}

@@

The tasks are executed by the node and pipe threads in parallel. The parent's
initialization methods are always executed before any child initialization
method. This parallelization is necessary to allow a speedy startup of the
configuration on large-scale graphics clusters. On the other hand, it means that
initialization functions are called even if the parent's initialization has
failed. \fig{fConfigInit} shows a sequence diagram of the config initialization.
\begin{figure}[ht!]\center
%\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
  {\caption{\label{fConfigInit}Config Initialization Sequence}}
\end{figure}

@@

{\footnotesize\lstinputlisting[firstline=310, lastline=317]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=319, lastline=327]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=329, lastline=359]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=387, lastline=390]
  {../../../src/examples/eqPly/config.cpp}}
{\footnotesize\lstinputlisting[firstline=450, lastline=481]
  {../../../src/examples/eqPly/config.cpp}}
{\footnotesize\lstinputlisting[firstline=528, lastline=534]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=284, lastline=308]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=790, lastline=859]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=257, lastline=268]
  {../../../src/examples/eqPly/config.cpp}}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\label{fDBDest}Destination View of a DB Compound using
      Demonstrative Coloring}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the frame data to ensure
that all channels render a given frame using the same position.

@@

{\footnotesize\lstinputlisting[firstline=591, lastline=610]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=619, lastline=689]
  {../../../src/examples/eqPly/channel.cpp}}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
\end{wrapfigure}
The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

For each window, one \textsf{GLXEventHandler} is allocated. FBO's and PBuffers
are handled in the same way as window drawables. The event dispatch finds the
corresponding event handler for each received event, which is then used to
process the event in a fashion similar to AGL. The \textsf{GLXWindowEvent}
passes the \textsf{XEvent} to \textsf{GLXWindowIF::\-process\-Event}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all their children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame}
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

\fig{fFrameSync} illustrates the local frame synchronization.
Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

@@

GLX initialization is very similar to AGL initialization. Again the steps are:
choose frame buffer configuration (pixel format), create OpenGL context and then
create drawable. The only difference is that the data returned by
\textsf{chooseGLXFBConfig} has to be freed using \textsf{XFree}:

{\footnotesize\lstinputlisting[firstline=67, lastline=101]
  {../../../src/libs/client/glXWindow.cpp}}

@@

\begin{wrapfigure}{r}{.612\textwidth}
  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
\end{wrapfigure}
The render clients can access a slave instance of the view using
\textsf{Channel::getView}. When called from one of the frame task
methods, this method will return the view of the current destination
channel for which the task method is executed. Otherwise it returns the
channel's native view, if it has one. Only destination channels of an
active canvas have a native view.

@@

The most common entity to subclass is the \textsf{View}, since the application
often amends with view-specific application data. A view might have a
distributable user data object, which has to inherit from \textsf{co::Object} to
be committed and synchronized with the associated object.

@@

\subsubsection{Run-time Layout Switch}

@@

\subsubsection{Run-time Stereo Switch}

Similar to switch a layout, the stereo mode of each view can be switched at
runtime. This causes all destination channels of the view to update the cyclop
eye or the left and right eye.

The configuration contains stereo information in three places: the view, the
segment and the compound.

The view defines which stereo mode is active: mono or stereo. This setting can
be done in the configuration file or programmatically.

The segment has the \textsf{eye} attribute in the configuration file which
defines which eyes are displayed by this segment. This is typically all eyes for
active stereo projections and the left or right eye for passive stereo. The
cyclop eye for passive stereo can either be set on a single segment or both,
depending if the second projector is active during monoscopic rendering. The
default setting enables all eyes for a segment.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/stereoSwitch.pdf}
  {\caption{\label{fStereoSwitch}Using two different decompositions during
      stereo and mono rendering}}
\end{wrapfigure}
The compound has an \textsf{eye} attribute defining which eyes it is capable of
updating. This allows to specify different compounds for the same channel,
depending on the stereo mode. One use case is to use one GPU each for updating a
stereo destination channel in stereo mode, and using a 2D compound with the two
GPUs in mono mode. \fig{fStereoSwitch} illustrates this example. The coloring in
mono mode is to illustrate the 2D decomposition. The default setting for the
compound eye attributes is all eyes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewUpdate.pdf}
  {\caption{\label{fViewUpdate}Event Flow during a View Update}}
\end{wrapfigure}
A view is derived from \textsf{eq::Frustum} (\fig{fViewUML}), and the
application process can set the wall or projection parameters at
runtime. For a description of wall and projection parameters please refer
to \sref{sFrustum}. The new data will be effective for the next frame.

The frustum of a view overrides the underlying frustum of the
segments. The frustum of a view does not typically correspond to a
physical projection, and can therefore be changed at runtime. The
frustum of a segment does however correspond to a physical projection,
and is therefore immutable.

@@

For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

\begin{wrapfigure}{l}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
Finding the correct assembly order for perspective frusta is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frusta. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.

@@

The eqPly example contains support for software anti-aliasing when being idle,
showing how to combine it with subpixel compounds. The same basic algorithm
might be applied to other multi-sampling effects, e.g., depth-of-field.

\subsubsection{Software Anti-Aliasing in eqPly}

@@

The subpixel area is a function of the current jitter step, the channel's
\textsf{subpixel} description and the idle state. To evenly distribute jitter
values for all channels over the total sub-pixel grid a precomputed prime number
table is used to pseudo-randomly select the subpixel:

{\footnotesize\lstinputlisting[firstline=547, lastline=570]
  {../../../src/examples/eqPly/channel.cpp}}

@@

@@

During idle rendering of any decomposition, the jitter for the frustum is
computed using the normalized subpixel center point and the size of a pixel on
the near plane. A random position within the sub-pixel is set as a sample
position, which will be used to move the frustum. The \textsf{getJitter} method
will return the the computed jitter vector for the current frustum. This method
has a default implementation in \textsf{eq::Channel} for subpixel compounds, but
is overwritten in \textsf{eqPly} to perform idle anti-aliasing:

{\footnotesize\lstinputlisting[firstline=491, lastline=533]
  {../../../src/examples/eqPly/channel.cpp}}

@@

Statistics are measured in milliseconds since the configuration was
initialized. The server synchronizes the per-configuration clock on each node
automatically. Each statistic event records the originator's (channel, window,
pipe, node, view or config) unique identifier.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/statistics}
  {\caption{\label{fStatistics}Statistics for a two node 2D compound}}
\end{wrapfigure}
Statistics are enabled per entity using an attribute hint. The hint determines
how precise the gathered statistics are. When set to \textsf{fastest}, the
per-frame clock is sampled directly when the event occurs. When set to
\textsf{nicest}, all OpenGL commands will be finished before sampling the
event. This may incur a performance penalty, but gives more correct results. The
default setting is fastest in release builds, and nicest in debug builds. The
fastest setting often attributes times to the operation causing an OpenGL
synchronization instead of the operation submitting the OpenGL commands, e.g.,
the readback time contains operations from the preceding draw operation.

@@

\subsubsection{\label{sStatisticsOverlay}Statistics Overlay}

The \textsf{eq::Channel} provides the method \textsf{drawStatistics} which
renders a statistics overlay using the gathered statistics events. Statistics
rendering is toggled on and off using the 's' key in the shipped Equalizer
examples.

\fig{fStatisticsDetail} shows a detailed view of \fig{fStatistics}. The
statistics shown are for a two-node 2D compound. The destination channel
is on the \textsf{appNode} and contributes to the rendering.

@@

The X axis is the time, the right-most pixel is the most current time. One
pixel on the screen corresponds to a given time unit, here one millisecond per
pixel. The scale is zoomed dynamically to powers-of-ten milliseconds to fit the
statistics into the available viewport. This allows easy and accurate
evaluations of bottlenecks or misconfigurations in the rendering pipeline. The
scale of the statistics is printed directly above the legend.

On the Y axis are the entities: channels, windows, nodes and the config. The
top-most channel is the local channel since it executes \textsf{frameAssemble},
and the lower channel is the remote channel, executing \textsf{frameReadback}.

To facilitate the understanding, older frames are gradually grayed out. The
rightmost, current frame is brighter than the frame before it.

The configuration uses the default latency of one frame. Consequently, the
execution of two frames overlaps. This can be observed in the early execution of
the remote channel's \textsf{frameDraw}, which starts while the local channel is
still drawing and assembling the previous frame.

The asynchronous execution allows operations to be pipelined, i.e., the
compression, network transfer and assembly with the actual rendering and
readback. This increases performance by minimizing idle and wait times. In this
example, the remote \textsf{channel2} has no idle times, and executes the
compression and network transfer of its output frame in parallel with rendering
and readback. Likewise, the application node receives and decompresses the frame
in parallel to its rendering thread.

In the above example, the local channel finishes drawing the frame early, and
therefore spends a couple of milliseconds waiting for the input frame from the
remote channel. These wait events, rendered red, are a sub-event of the yellow
\textsf{frameAssemble} task. Using a \textsf{load\_equalizer} instead of a
static 2D decomposition would balance the rendering in this example better and
minimize or even eliminate this wait time.

The white \textsf{Window::swapBuffers} task might take a longer time, since the
execution of the swap buffer is locked to the vertical retrace of the
display. Note that the remote source window does not execute swapBuffers in
this configuration, since it is a single-buffered FBO.

The beginning of a frame is marked by a vertical green line, and the end of a
frame by a vertical gray line. These lines are also attenuated. The brightness
and color matches the event for \textsf{Config::startFrame} and
\textsf{Config::finishFrame}, respectively. The event for \textsf{startFrame} is
typically not visible, since it takes less than one millisecond to execute. If
no idle processing is done by the application, the event for
\textsf{finishFrame} occupies a full frame, since the config is blocked here
waiting for the frame \textsf{current - latency} to complete.

A legend directly below the statistics facilitates understanding. It lists the
per-entity operations with its associated color. Furthermore, some other textual
information is overlayed with the statistics. The total compression ratio is
printed with each readback and compression statistic. In this case the image has
been not been compressed during download. For network transfer it has been
compressed to 1\% of its original size since it contains a lot of black
background pixels. For readback the plugin with the name \textsf{0x101,
  EQ\_COMPRESSOR\_TRANS\-FER\_RGBA\_TO\_BGRA} has been used, and for compression
the plugin \textsf{0x11, EQ\_COM\-PRESSOR\_RLE\_DIFF\_BGRA} was used. The
compressor names are listed in the associated plugin header
\textsf{compressorTypes.h}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
  {\caption{\label{fNetNode}Communication between two Nodes}}
\end{wrapfigure}
\fig{fNetNode} shows the communication between two nodes. The nodes in
the listening state also have a listening connection, which is not shown
in the Figure for complexity reasons.

When the remote node sends a packet, the listening node receives the
packet and dispatches it from the receiver thread using the method
\textsf{dispatchCommand}. The default implementation knows how to
dispatch packets of type node, session or object. Applications can
define custom data types for packets, and then have to extend
\textsf{dispatchCommand} to handle these custom data types.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
  {\caption{\label{fNetObject}Communication between two Objects}}
\end{wrapfigure}
Multiple instances of the same object (identifier) can be mapped on the same
node and session simultaneously. Each object instance has a node-unique instance
identifier. When the instance ID is set to \textsf{EQ\_INSTANCE\_ALL} (the
default value), a command is dispatch to all object instances registered to the
session with the correct object ID. Multiple object instance per node are used
when different threads on the node need to process a different version of the
same object.

When the packet is intended for a specific instance, set the instance identifier
of the object packet to the corresponding value.
